---
title: "Survey of Consumer Finance student loan forgiveness analysis by groups--student proposals"
author: "Harold Pollack"
date: "1/28/2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, echo=FALSE, warn=-1}
library("UsingR")
library("dplyr")
library("knitr")
library("ggplot2")
library("xts")
library("questionr")
library("janitor")
library("lubridate")
library("epiDisplay")
library("reshape2")
library("plotly")
library("formattable")
library("data.table")
library("naniar") 
library("scales") 
options(width = 1200)
knitr::opts_chunk$set(echo = TRUE)
```

This markdown file analyzes student loan data 2019 Survey of Consumer Finance files.
In particular, it models proposals to "forgive" the first $K owed within defined populations, defined by an indicator dselect=1. A function does the rest of the work.
It provides the basis for some policies for student loan forgiveness.
We only examine the first six loans held by respondent or spouse.
We also follow Gale, Gelfond, Fitchtner, and Harris (2020) by multiplying SCF’s survey weight x42001 by 0.2 to produce summary statistics.
https://protectborrowers.org/new-data-show-borrowers-of-color-and-low-income-borrowers-are-missing-out-on-key-protections-raising-significant-fair-lending-concerns/
We will restrict to loans owed by respondent or spouse, given data challenges to evaluating loans by other household members





```{r read SCF 2019, cache=TRUE, echo=FALSE}
#
#    Read 2019 Survey of Consumer finance
#    Original dataset downloaded from https://www.federalreserve.gov/econres/files/scf2019ascii.zip
#    The dataframe will be used in the analysis
#
load("/Users/haroldpollack/Documents/SCF/p19i6.rdata")
df_student_loan<-p19i6
```

Forgiveness with distribution across groups

```{r function for descriptives on forgiveness across groups, echo=FALSE}
forgive_descript_groups<-function(df, forgive_threshold)
#
#      This provides more detail on where the money is going.
#
{
 #
 #   df is the basic student loan dataframe, deselect will be the key variable to select populations eligible for forgiveness
 #   forgive_threshold is the level of loans to be forgiven within the population.
 #
  print(" ")
  print( paste("forgiveness threshold--",prettyNum(forgive_threshold,big.mark = ",")))   
  print(" ")
  #
  #   Somewhat wastefully, I create this dataframe df2 that includes the variables we will use for the analysis.
  #
  df2<-subset(df,dselect==1,select=c(dblack,dhisp,x42001,total_loan_amount,dselect,dpublic_aid,fam_income,total_loan_amount_owed) )
  thresh_print<-prettyNum(forgive_threshold,big.mark = ",")
  #
  # We compute the weighted amount owed--Note that the SCF is a weighted sample, with weight x42001 in the dataset
  #
  Mean_loan_amount_owed <- prettyNum(weighted.mean(df2$total_loan_amount_owed,df2$x42001),big.mark = ",")
  print( paste("Mean loan amount borrowed in the group with threshold ",thresh_print, " ",Mean_loan_amount_owed)) 
  #
  #    We compute the amount of money forgiven if we forgive up to the first $forgive_threshold of what people owe.
  #
  #    pmin is an annoying nicety required by R to create a variable that indicates the amount of loan forgiven.
  #
  df2 <- transform(df2, amount_forgiven = pmin(total_loan_amount_owed, forgive_threshold))
  #
  #      Now we compute the mean amount forgiven
  #
  Mean_loan_amount_forgiven <- prettyNum(weighted.mean(df2$amount_forgiven,df2$x42001),big.mark = ",")
  print( paste("Mean loan amount forgiven in the group with threshold ",thresh_print, " ",Mean_loan_amount_forgiven)) 
  #
  #       This is how much people owe after forgiveness
  #
 df2$post_forgiveness_amount_owed<-df2$total_loan_amount_owed-df2$amount_forgiven
  #
  #      Now we add up across the whole sample to estimate the total number of dollars forgiven. The sum is weighted by each observation's sample weight
  #
 total_loan_amount_forgiven <- prettyNum( 0.2*sum(df2$amount_forgiven*df2$x42001),big.mark = ",")
    print( paste("Estimated total loan amount forgiven  ","$",total_loan_amount_forgiven)) 
#
# Now provide total amount forgiven by key groups of interest dblack, dhisp, and so on are dummy 0-1 variables.
#
#       Total amount owed by Black respondents
#       I create a dummy (0-1) variable for people in the category.
#
total_loan_amount_forgiven_black <- prettyNum( 0.2*sum(df2$amount_forgiven*df2$x42001*df2$dblack),big.mark = ",")
    print( paste("Estimated total loan amount forgiven--Black group  ","$",total_loan_amount_forgiven_black))
#
#       Total amount owed by Hispanic respondents
#       I create a dummy (0-1) variable for people in the category.
#
total_loan_amount_forgiven_hisp <- prettyNum( 0.2*sum(df2$amount_forgiven*df2$x42001*df2$dhisp),big.mark = ",")
    print( paste("Estimated total loan amount forgiven--Hispanic group  ","$",total_loan_amount_forgiven_hisp)) 
#
#      How much of the money goes to roughly the bottom quarter of the family income distribution--about $34,000/yr. Note that there are some negative incomes.
#      I create a dummy (0-1) variable for people in the category 
#
df2$d34000<-as.numeric(df2$fam_income>=-900000 & df2$fam_income<=34000)
total_loan_amount_forgiven_34000 <- prettyNum( 0.2*sum(df2$amount_forgiven*df2$x42001*df2$d34000),big.mark = ",")
    print( paste("Estimated total loan amount forgiven--Family income less than $34,000  ","$",total_loan_amount_forgiven_34000)) 
#
#      How much of the money goes to families with SNAP, SSI, TANF or other public assistance income? .
#
total_loan_amount_forgiven_public_aid <- prettyNum( 0.2*sum(df2$amount_forgiven*df2$x42001*df2$dpublic_aid),big.mark = ",")
    print( paste("Estimated total loan amount forgiven--Families receiving public aid income (e.g. SSI/TANF/SNAP)  ","$",total_loan_amount_forgiven_public_aid )) 
#
#      How much of the money goes to high-income families with incomes exceeding $100,000/yr. 
#
df2$d100000<-as.numeric(df2$fam_income>=100000 & df2$fam_income<=100000000)
#
total_loan_amount_forgiven_100000 <- prettyNum( 0.2*sum(df2$amount_forgiven*df2$x42001*df2$d100000),big.mark = ",")
    print( paste("Estimated total loan amount forgiven--Family income greater than $100,000  ","$",total_loan_amount_forgiven_100000)) 
#
#      Dummy variable for loan forgiveness, then weighted by the sample weight
#
 df2$dforgiveness<-as.numeric(df2$amount_forgiven>=1 & df2$amount_forgiven<100000000)
 df2$wt_forgiven<-df2$dforgiveness*df2$x42001
 #
 #    Find the distribution of family income among those with loan forgiveness. I only use the 25th, 50th, and 75th percentile
 #
v_owed<- wtd.quantile (df2$fam_income, probs=c(0, 0.2, 0.25, 0.3, 0.4, 0.5, 0.6, 0.7, 0.75, 0.8, 0.85,0.9,0.95, 0.99, 1), na.rm = FALSE, weight=df2$wt_forgiven)
print(paste("25th percentile of Family income distribution in this forgiven group when amount owed below $", thresh_print, "--",prettyNum(as.numeric(v_owed)[3],big.mark = ",")))
print(paste("50th percentile of Family income distribution in this forgiven group when amount owed below $", thresh_print, "--",prettyNum(as.numeric(v_owed)[6],big.mark = ",")))
print(paste("75th percentile of Family income distribution in this forgiven group when amount owed below $", thresh_print, "--",prettyNum(as.numeric(v_owed)[9],big.mark = ",")))
#
#   Proportion of the demographic groups with at least some loan forgiveness in each group.
#
Mean_black_forgiven <- weighted.mean(df2$dblack,df2$wt_forgiven)
  print( paste("Mean proportion Black--forgiven debt below $",thresh_print, " ", Mean_black_forgiven)) 
Mean_hisp_forgiven <- weighted.mean(df2$dhisp,df2$wt_forgiven)
  print( paste("Mean proportion Hisp--forgiven debt below $",thresh_print, " ", Mean_hisp_forgiven)) 

Mean_public_aid_forgiven <- weighted.mean(df2$dpublic_aid,df2$wt_forgiven)
print( paste("Mean proportion Public aid--forgiven debt below $",thresh_print, " ", Mean_public_aid_forgiven)) 
  v<-c(Mean_black_forgiven,Mean_hisp_forgiven,Mean_public_aid_forgiven)
return(v)
}
```

Now we get to the program.


```{r manipulate new SCF file,echo=FALSE}
#
#   Show some characteristics of the SCF data
#
#    Number of obs and other characteristics
#
nrow2<-nrow(df_student_loan)
print( paste("Number of observations: ", nrow2))
summary(df_student_loan$x7179)
print( paste("Student loan?: ")) 
table(df_student_loan$x7801)
print( paste("Number of loans: ")) 
table(df_student_loan$x7802)
sumwt<-sum(df_student_loan$x42001)
print( paste("Sum of weights--full sample: ", sumwt))

#
#     Keep only the group with loans     
#
df_student_loan <- subset(df_student_loan, x7802 > 0 & x7802 < 8) 


#
#    Define key variables
#    Public assistance x5719=1
#
df_student_loan$dpublic_aid<-as.numeric(df_student_loan$x5719==1)
print( paste("Public assistance income? ")) 
table(df_student_loan$dpublic_aid)
#
# Does loan belong to householder or spouse
#
df_student_loan$dloan_own1<-as.numeric(df_student_loan$x7978==1|df_student_loan$x7978==2)
print( paste("Loan1 for student or spouse? ")) 
table(df_student_loan$dloan_own1,df_student_loan$x7802)


df_student_loan$dloan_own2<-as.numeric(df_student_loan$x7883==1|df_student_loan$x7883==2) 
print( paste("Loan2 for student or spouse? ")) 
table(df_student_loan$dloan_own2,df_student_loan$x7802)

df_student_loan$dloan_own3<-as.numeric(df_student_loan$x7888==1|df_student_loan$x7888==2) 
print( paste("Loan3 for student or spouse? ")) 
table(df_student_loan$dloan_own3,df_student_loan$x7802)

df_student_loan$dloan_own4<-as.numeric(df_student_loan$x7893==1|df_student_loan$x7893==2) 
print( paste("Loan4 for student or spouse? ")) 
table(df_student_loan$dloan_own4,df_student_loan$x7802)

df_student_loan$dloan_own5<-as.numeric(df_student_loan$x7898==1|df_student_loan$x7898==2) 
print( paste("Loan5 for student or spouse? ")) 
table(df_student_loan$dloan_own5,df_student_loan$x7802)

df_student_loan$dloan_own6<-as.numeric(df_student_loan$x7993==1|df_student_loan$x7993==2) 
print( paste("Loan6 for student or spouse? ")) 
table(df_student_loan$dloan_own6,df_student_loan$x7802)

```


```{r, manipulate loan totals,echo=FALSE}

sumwt<-sum(df_student_loan$x42001)
#summary(df_student_loan$x42001)
#
#     Age
#
#print( paste("Age variable:"))
#table(df_student_loan$x8022)
#print( paste("Age variable:"))
#table(df_student_loan$x13)
print( paste("Hispanic/Latino:"))
table(df_student_loan$x7004)
print( paste("race variable:"))
table(df_student_loan$x6809)
print( paste("Sum of weights student loan sample: ", sumwt))
#
#    Race ethicity
#
df_student_loan$dblack<-as.numeric(df_student_loan$x6809==2|df_student_loan$x6810==2)
#
#     Hispanic/Latino defined as x7004=1. Not mutually exclusive with black or other race variables.
#
df_student_loan$dhisp<-as.numeric(df_student_loan$x7004==1)
df_student_loan$drace_other<-as.numeric(df_student_loan$x6809==-7 & df_student_loan$dblack==0 & df_student_loan$dhisp==0 )
print( paste("Black by Hispanic table"))
table(df_student_loan$dblack,df_student_loan$dhisp)
table(df_student_loan$x6809,df_student_loan$dhisp)
table(df_student_loan$x6810,df_student_loan$dhisp)

print( paste("Other category"))
table(df_student_loan$drace_other)
#
#      Let's restrict to loans owed by householder or spouse--difficult to analyze loans held by children or others in the family/household
#
df_student_loan$total_loan_amount<-df_student_loan$x7805*df_student_loan$dloan_own1+df_student_loan$x7828*df_student_loan$dloan_own2+df_student_loan$x7851*df_student_loan$dloan_own3+df_student_loan$x7905*df_student_loan$dloan_own4+df_student_loan$x7928*df_student_loan$dloan_own5+df_student_loan$x7951*df_student_loan$dloan_own6
print( paste("Total loan amount--householder and spouse only")) 
quantile(df_student_loan$total_loan_amount, probs = seq(0, 1, by= 0.05)) # decile
wtd.quantile (df_student_loan$total_loan_amount, probs=c(0, 0.2, 0.25, 0.3, 0.4, 0.5, 0.6, 0.7, 0.75, 0.8, 0.85,0.9,0.95, 0.99, 1), na.rm = FALSE, weight=df_student_loan$x42001)

#
#    Public assistance
#
df_student_loan$wtp_ass<-df_student_loan$x42001*df_student_loan$dpublic_aid
#




df_student_loan$dstudent_loan<-as.numeric((df_student_loan$x7801==1))
#print( paste("Student loans"))
#table(df_student_loan$dstudent_loan)
df_student_loan$fam_income<-df_student_loan$x5729
fam_income_overall<- weighted.mean(df_student_loan$fam_income, df_student_loan$x42001)
print( paste("Mean family income--overall: ", fam_income_overall))
print( paste("Family income")) 
quantile(df_student_loan$total_loan_amount, probs = seq(0, 1, by= 0.05)) # decile
df_student_loan$wt_student_loan<- df_student_loan$x42001*df_student_loan$dstudent_loan
print( paste("Student loan weight: ")) 
summary(df_student_loan$wt_student_loan)       
df_student_loan$wt_no_student_loan<-df_student_loan$x42001-df_student_loan$wt_student_loan
#
#   Suppressed the no student loan data in this iteration
#
#print( paste("No student loan weight: ")) 
#summary(df_student_loan$wt_no_student_loan)       
fam_income_loan<- weighted.mean(df_student_loan$fam_income, df_student_loan$wt_student_loan)
#fam_income_no_loan <- weighted.mean(df_student_loan$fam_income, df_student_loan$wt_no_student_loan)
print( paste("Mean family income--loans: ", fam_income_loan))
#print( paste("Mean family income--no loans: ", fam_income_no_loan))

```

Now amount still owed

```{r amount still owed, echo=FALSE}
#
#    total loan amount owed
#
df_student_loan$total_loan_amount_owed<-df_student_loan$x7824*df_student_loan$dloan_own1+df_student_loan$x7847*df_student_loan$dloan_own2+df_student_loan$x7870*df_student_loan$dloan_own3+df_student_loan$x7924*df_student_loan$dloan_own4+df_student_loan$x7947*df_student_loan$dloan_own5+df_student_loan$x7970*df_student_loan$dloan_own6
print( paste("Total loan amount still owed")) 
summary(df_student_loan$total_loan_amount_owed)
wtd.quantile (df_student_loan$total_loan_amount_owed, probs=c(0, 0.2,  0.25, 0.3, 0.4, 0.5, 0.6, 0.7, 0.75, 0.8, 0.85,0.9,0.95, 0.99, 1), na.rm = FALSE, weight=df_student_loan$wt_student_loan)


df_student_loan$dstudent_loan<-as.numeric((df_student_loan$x7801==1))
#print( paste("Student loans still owed"))
table(df_student_loan$dstudent_loan)
df_student_loan$fam_income<-df_student_loan$x5729
fam_income_overall<- weighted.mean(df_student_loan$fam_income, df_student_loan$x42001)
print( paste("Mean family income--overall: ", fam_income_overall))

df_student_loan$wt_student_loan<- df_student_loan$x42001*df_student_loan$dstudent_loan
df_student_loan$wt_no_student_loan<-df_student_loan$x42001-df_student_loan$wt_student_loan
fam_income_loan<- weighted.mean(df_student_loan$fam_income, df_student_loan$wt_student_loan)

print( paste("Mean family income--loans owed: ", fam_income_loan))
#print( paste("Mean family income--no loans owed: ", fam_income_no_loan))

Mean_loan_amount <- weighted.mean(df_student_loan$total_loan_amount,df_student_loan$wt_student_loan)
print( paste("Mean loan amount owed: ", Mean_loan_amount))
```

Looping through forgiveness outcomes, with calls to forgive_descript_groups given different selected groups and levels of forgiveness

```{r, looping through the forgiveness function,echo=FALSE}
# 
# Sanders proposals--Basically everyone. Top-code at $1 million for data reasons
#
df_student_loan$dselect<- 1
thresh_label<-"All loans less than $1 million"
print( paste("All loans less than $1 million"))
forgive_threshold<-1000000
print( paste("forgiveness threshold ",prettyNum(forgive_threshold,big.mark = ",")))  
forgive_descript_groups(df_student_loan,forgive_threshold)

# 
# All black and Hispanic with loans less than $1,000,000
#
df_student_loan$dselect<- as.numeric( df_student_loan$dblack==1 | df_student_loan$dhisp==1)
thresh_label<-"Black or Hispanic with loans less than $1million"
print( paste("Black or Hispanic"))
forgive_threshold<-1000000
print( paste("forgiveness threshold ",prettyNum(forgive_threshold,big.mark = ",")))  
forgive_descript_groups(df_student_loan,forgive_threshold)
# 
# All households receiving public assistance income
#
df_student_loan$dselect<- as.numeric( df_student_loan$dpublic_aid==1)
thresh_label<-"Public aid income with loans less than $1million"
print( paste("Public aid"))
forgive_threshold<-1000000
print( paste("forgiveness threshold ",prettyNum(forgive_threshold,big.mark = ",")))  
forgive_descript_groups(df_student_loan,forgive_threshold)
# 
# Forgive first $50,000 for all families with incomes less than $34,000
#  $34,000 is about the 25th percentile of income
#
df_student_loan$dselect<- as.numeric( df_student_loan$fam_income<=34000)
thresh_label<-"First $50,000 for all families with incomes less than $34,000"
print( paste("Forgive first $50,000 for all families with incomes less than $34,000"))
forgive_threshold<-50000
print( paste("forgiveness threshold ",prettyNum(forgive_threshold,big.mark = ",")))  
forgive_descript_groups(df_student_loan,forgive_threshold)

# 
# Forgive first $100,000 when family income is less than $25000 and there is public assistance income
#
df_student_loan$dselect<- as.numeric(df_student_loan$fam_income<=25000 & df_student_loan$dpublic_aid==1)
thresh_label<-"Family income less than $25000"
print( paste("Forgive first $100,000 when family income is less than $25,000 and there is public assistance income, "))
forgive_threshold<-100000
print( paste("forgiveness threshold ",prettyNum(forgive_threshold,big.mark = ",")))  
forgive_descript_groups(df_student_loan,forgive_threshold)

#
#     Now just do thresholds
#

# 
# Forgive first $10,000 
#
df_student_loan$dselect<- 1
thresh_label<-"Forgive first $10,000 for all families"
print( paste("Forgive first $10,000 for all families, "))
forgive_threshold<-10000
print( paste("forgiveness threshold ",prettyNum(forgive_threshold,big.mark = ",")))  
forgive_descript_groups(df_student_loan,forgive_threshold)

# 
# Forgive first $25,000 
#
df_student_loan$dselect<- 1
thresh_label<-"Forgive first $25,000 for all families"
print( paste("Forgive first $25,000 for all families, "))
forgive_threshold<-25000
print( paste("forgiveness threshold ",prettyNum(forgive_threshold,big.mark = ",")))  
forgive_descript_groups(df_student_loan,forgive_threshold)

# 
# Forgive first $50,000 
#
df_student_loan$dselect<- 1
thresh_label<-"Forgive first $50,000 for all families"
print( paste("Forgive first $50,000 for all families, "))
forgive_threshold<-50000
print( paste("forgiveness threshold ",prettyNum(forgive_threshold,big.mark = ",")))  
forgive_descript_groups(df_student_loan,forgive_threshold)

# 
# Forgive first $75,000 
#
df_student_loan$dselect<- 1
thresh_label<-"Forgive first $75,000 for all families"
print( paste("Forgive first $75,000 for all families, "))
forgive_threshold<-75000
print( paste("forgiveness threshold ",prettyNum(forgive_threshold,big.mark = ",")))  
forgive_descript_groups(df_student_loan,forgive_threshold)

# 
# Forgive first $100,000 
#
df_student_loan$dselect<- 1
thresh_label<-"Forgive first $100,000 for all families"
print( paste("Forgive first $100,000 for all families, "))
forgive_threshold<-100000
print( paste("forgiveness threshold ",prettyNum(forgive_threshold,big.mark = ",")))  
forgive_descript_groups(df_student_loan,forgive_threshold)


```